---
title: "Wine Prediction"
author: "Samson Akomolafe"
date: "2024-11-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#To carry out wine quality prediction on a train dataset, I am going to follow these enumerated steps below:

#1.    Data Preparation

##1a.    Loading the Data Set: Import necessary libraries and load training dataset.
      
##1b.    Explore the Data: Understand the dataset structure, check for missing values, and examine the target variable (wine quality).
      
##1c.    Data Cleaning: Handle any missing values and remove duplicates if necessary.

#2.    Data Pre-processing

##2a.    Feature Selection: Identify and select relevant features that may affect wine         quality.

##2b.    Normalization/Standardization: Scale the features to ensure they are on a similar scale, which can be essential for some algorithms.(if required)

##2c.    Encoding Categorical Variables: If there are any categorical features, convert them into numerical format using techniques like one-hot encoding.

#3.     Splitting the Data

##3a.    If you have a validation set, split the training data into training and validation sets. This can be done using train_test_split from sklearn.

#4. Model Selection

##4a.    Choose a machine learning model appropriate for the task. Common algorithms for regression or classification tasks include:

##4b    For this Project, I am going to use these 2 regression model and select the model that predicted better for the testing data set. These models are;
##4b1.    Decision Trees
##4b2.    Multiple Linear Regression

#5.       Model Training

##5a.   Fit the model to the training data using the selected features and target variable.

#6.     Model Evaluation

##6a.   Evaluate the model's performance using appropriate metrics:

##6b.   For regression: I am going to employ Mean Squared Error (MSE), R² score. And for classification I will use Accuracy, and Recall.

#7.   Hyperparameter Tuning- I am not going to do hyperparameter tuning for this modelling.

#8. Make Predictions: I am going to use the trained model to make quality predictions on the provided test dataset.

```{r}

#1.    Data Preparation

#1a.    Loading the Data Set: Import necessary libraries and load training dataset.
      
#1b.    Explore the Data: Understand the dataset structure, check for missing values, and examine the target variable (wine quality).

# Load required libraries
library(caret)
library(randomForest)
library(dplyr)
library(reshape2)
library(corrplot)
library(GGally)
library(plotly)

# Load the train dataset
wine_data <- read.csv(file.choose(),header = TRUE)


# Explore the dataset
str(wine_data)

#observation
#1. The train dataset has 15 variables
#2. There are 2 categorical variables in the data, location and type
#3. The data has 5463 columns
#4. The dataset has integer, string and numerical variables
#5. The dataset has a quality column, that is, the prediction variable in the dataset

#Understand the structure of the dataset
colSums(is.na(wine_data))  # Count missing values in each column

#Observation
#1. There are no missing values in the dataset
#2. I will proceed to the next step of my work.

#EXPLORATIVE DATA ANALYSIS *EDA*
#Indept work on the EDA section have been presented as a group exercise with Hayoung #and Jack.However, below is some of the codes we used for the presentations.

#Quality
ggplot(wine_data, aes(x = quality)) + 
  geom_bar(fill = "royalblue") + 
  labs(title = "Distribution of Wine Quality", x = "Quality", y = "Count")
#Please refer to my EDA group presentation for observation

#Alcohol Content
ggplot(wine_data, aes(x = alcohol)) + 
  geom_histogram(bins = 30, fill = "lightgreen", color = "black") + 
  labs(title = "Distribution of Alcohol Content", x = "Alcohol", y = "Frequency")
#Please refer to my EDA group presentation for observation

#pH Levels
ggplot(wine_data, aes(x = pH)) + 
  geom_histogram(bins = 30, fill = "lightcoral", color = "black") + 
  labs(title = "Distribution of pH Levels", x = "pH", y = "Frequency")
#Please refer to my EDA group presentation for observation

#Citrate Levels
ggplot(wine_data, aes(x = citric.acid)) + 
  geom_histogram(bins = 30, fill = "gold", color = "black") + 
  labs(title = "Distribution of Citric Acid Levels", x = "Citric Acid", y = "Frequency")
#Please refer to my EDA group presentation for observation

#Quality vs Alcohol
ggplot(wine_data, aes(x = alcohol, y = quality)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE, color = "red") + 
  labs(title = "Quality vs Alcohol", x = "Alcohol", y = "Quality")
#Please refer to my EDA group presentation for observation
     
#pH vs Quality
ggplot(wine_data, aes(x = pH, y = quality)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE, color = "blue") + 
  labs(title = "Quality vs pH", x = "pH", y = "Quality")
#Please refer to my EDA group presentation for observation
     
#Citrate Levels vs Quality:
ggplot(wine_data, aes(x = citric.acid, y = quality)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE, color = "green") + 
  labs(title = "Quality vs Citric Acid", x = "Citric Acid", y = "Quality")
#Please refer to my EDA group presentation for observation
     
# Select only numeric columns
numeric_wine_data <- wine_data %>%
  select(where(is.numeric))
correlation_matrix <- cor(numeric_wine_data)
melted_cormat <- melt(correlation_matrix)
ggplot(data = melted_cormat, aes(Var1, Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(limit = c(-1, 1), low = "red", high = "green", mid = "white", 
                          midpoint = 0, space = "Lab", name="Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, size = 12, hjust = 1)) +
  coord_fixed()
#Please refer to my EDA group presentation for observation

#Box plots for categorical analysis:
#To analyze the distribution of alcohol or pH levels across different quality levels:
ggplot(wine_data, aes(x = as.factor(quality), y = alcohol)) + 
  geom_boxplot(fill = "lightblue") + 
  labs(title = "Alcohol Content by Quality", x = "Quality", y = "Alcohol")
#Please refer to my EDA group presentation for observation

ggplot(wine_data, aes(x = as.factor(quality), y = citric.acid)) + 
  geom_boxplot(fill = "green") + 
  labs(title = "Alcohol Content by Quality", x = "Quality", y = "Citric Acid")
#Please refer to my EDA group presentation for observation

# Summarize average quality by location and type
summary_data <- wine_data %>%
  group_by(location, type) %>%
  summarise(avg_quality = mean(quality, na.rm = TRUE),
            count = n()) %>%
  arrange(desc(avg_quality))
#Please refer to my EDA group presentation for observation

# View the summary data
print(summary_data)


# Create a bar plot
ggplot(summary_data, aes(x = reorder(location, -avg_quality), y = avg_quality, fill = type)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Average Wine Quality by Location and Type",
       x = "Location",
       y = "Average Quality") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
#Please refer to my EDA group presentation for observation


# ANOVA test
anova_result <- aov(quality ~ location * type, data = wine_data)
summary(anova_result)
#Please refer to my EDA group presentation for observation

# Convert 'type' and 'location' to factors
wine_data$type <- as.factor(wine_data$type)
wine_data$location <- as.factor(wine_data$location)

# Create a pairplot for quality, type, and location
ggpairs(wine_data, 
        columns = c("quality", "type", "location"), 
        mapping = aes(color = type), 
        lower = list(continuous = wrap("points", alpha = 0.6)),
        upper = list(continuous = wrap("cor", size = 3))) +
        labs(title = "Wine Quality by Location and Type")
#Please refer to my EDA group presentation for observation

#quick view of the data
table(wine_data$quality, wine_data$location, wine_data$type)
table(wine_data$quality)

#3D plptting of wine data
plot_ly(data = wine_data, 
        x = ~volatile.acidity, 
        y = ~density,
        z = ~alcohol, 
        type = "scatter3d", 
        mode = "markers",
        color = ~as.factor(quality), 
        marker = list(size = 4,  opacity = 0.8)) |>
  layout(title = "3D Scatterplot: Location of Wine by Alcohol, Density, and Volatile Acidity",
         scene = list(xaxis = list(title = "Volatile Acidity"),
                      yaxis = list(title = "Density"),
                      zaxis = list(title = "Alcohol")),
         legend = list(title = list(text = "Legend")))
#Please refer to my EDA group presentation for observation

################################
#END OF EDA SECTION
################################

```

```{r}


###################################################
#MODEL 1 - LINEAR REGRESSION BY DECISION TREE MODEL
###################################################
   
#Pre-Processing of Data for modelling
#Separate the Categorical Variables
numerical_features <- names(wine_data)[sapply(wine_data, is.numeric)] 
categorical_features <- wine_data[, c("type", "location")]

# Create dummy variables for the specified categorical features
dummy_vars <- model.matrix(~ type + location - 1, data = wine_data)

# Combine dummy variables with the original numerical features
wine_data_with_dummies <- cbind(wine_data[numerical_features], dummy_vars)

# View the resulting data frame
head(wine_data_with_dummies)

# Split the dataset into training and testing sets
set.seed(123)
train_index <- createDataPartition(wine_data_with_dummies$quality, p = 0.8, list = FALSE)
train_data <- wine_data_with_dummies[train_index, ]
test_data <- wine_data_with_dummies[-train_index, ]

# Train a random forest regression model
rf_model <- randomForest(quality ~ . - ID, data = train_data, importance = TRUE, ntree = 1000)

# Print the model summary
print(rf_model)

# Analyze feature importance
importance_values <- importance(rf_model)
importance_df <- data.frame(Feature = rownames(importance_values), Importance = importance_values[, 1])
importance_df <- importance_df %>%
  arrange(desc(Importance))

# Print feature importance
print(importance_df)

# Plot feature importance
barplot(importance_df$Importance, names.arg = importance_df$Feature, las = 2, 
        main = "Feature Importance", col = "blue", cex.names = 0.7)

# Predict on the test Data set
test_dat <- read.csv(file.choose(), header=TRUE)

#Prepare the test dataset to match the structure train data
#Separate the Categorical Variables
num_features <- names(test_dat)[sapply(test_dat, is.numeric)] 
cat_features <- test_dat[, c("type", "location")]

# Create dummy variables for the specified categorical features
dummy_vars_test <- model.matrix(~ type + location - 1, data = test_dat)

# Combine with numerical features if needed
test_data_rf <- cbind(test_dat[num_features], dummy_vars_test)

#Predict quality on test data
Quality <- round(predict(rf_model, newdata = test_data_rf))

#Combine predicted "Wine Quality" column with test data
New_wine_data = data.frame(test_data_rf, Quality)

#Display the predicted test data
print(New_wine_data)

# Evaluate the model performance
results_rf <- data.frame(Actual = New_wine_data$Quality, Predicted = Quality)
mse_rf <- mean((results_rf$Actual - results_rf$Predicted)^2)
rmse_rf <- sqrt(mse_rf)

# Print the evaluation metrics
cat("Root Mean Squared Error (RMSE):", rmse_rf, "\n")

# Optionally, I will plot actual vs predicted values
plot(results_rf$Actual, results_rf$Predicted, xlab = "Actual Quality", ylab = "Predicted Quality", 
     main = "Actual vs Predicted Wine Quality", pch = 19, col = "blue")
abline(0, 1, col = "red")

# Create confusion matrix
confusion_mat_rf <- confusionMatrix(as.factor(results_rf$Predicted), as.factor(results_rf$Actual))
print(confusion_mat_rf)

# Create a confusion matrix table
cm_table_rf <- as.data.frame(confusion_mat_rf$table)

# Plot confusion matrix
ggplot(data = cm_table_rf, aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "blue") +
  geom_text(aes(label = Freq), vjust = 1) +
  labs(title = "Confusion Matrix", x = "Actual Quality", y = "Predicted Quality") +
  theme_minimal()

# Extract TN, FP, FN, TP
TN <- confusion_mat_rf$table[1, 1]
TP <- confusion_mat_rf$table[2, 2]
FP <- confusion_mat_rf$table[1, 2]
FN <- confusion_mat_rf$table[2, 1]

# Print the values
cat("True Negatives (TN):", TN, "\n")
cat("True Positives (TP):", TP, "\n")
cat("False Positives (FP):", FP, "\n")
cat("False Negatives (FN):", FN, "\n")

# Create a data frame for plotting confusion matrix values
cm_values_rf <- data.frame(
  Type = c("True Negatives", "False Positives", "False Negatives", "True Positives"),
  Count = c(TN, FP, FN, TP)
)

# Plot confusion matrix values
ggplot(cm_values_rf, aes(x = Type, y = Count, fill = Type)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("yellow", "blue", "orange", "skyblue")) +
  labs(title = "Confusion Matrix Values", x = "Type", y = "Count") +
  theme_minimal() +
  theme(legend.position = "none")


```

#I wil now build multiple linear regression to ne able to compare both models and then choose the one.

```{r}

###################################################
#MODEL 2 - MULTIPLE LINEAR REGRESSION 
###################################################

# Load necessary packages
library(tidyr)
library(ggplot2)


# Load the dataset
df <- read.csv(file.choose(),header = TRUE)


table(df$quality,df$location)

table(df$type,df$location)

#Separate the Categorical Variables
nume_features <- names(df)[sapply(df, is.numeric)] 
cate_features <- df[, c("type", "location")]

# Create dummy variables for the specified categorical features
dummies_df <-  model.matrix(~ type + location - 1, data = df)

# Combine with numerical features if needed
wine_encoded <- cbind(df[, nume_features], dummies_df)


#Model Building
# For reproducibility
set.seed(123)  
train_index_lr <- sample(1:nrow(wine_encoded), 0.7 * nrow(df))
train_data_lr <- wine_encoded[train_index, ]
test_data_lr <- wine_encoded[-train_index, ]


# Fit the model
model_lr <- lm(quality ~ ., data = train_data_lr)

# Summary of the model
summary(model_lr)

#Save the model in RDS file format
#saveRDS(model_lr, "wine_trained_model.rds")
#this file is needed, I will use it as my Shiny app trained model to be able to predict wine quality on the test data

# Get the coefficients
importance <- summary(model_lr)$coefficients
importance <- importance[-1, ]  # Remove the intercept
importance <- importance[order(-abs(importance[, 1])), ]  # Sort by absolute value
print(importance)

# Prepare the data for plotting
importance_df <- data.frame(Variable = rownames(importance),
                            Coefficient = importance[, 1])

# Plotting the bar chart
ggplot(importance_df, aes(x = reorder(Variable, abs(Coefficient)), y = Coefficient)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +  # Flip the axes for better readability
  labs(title = "Variable Importance from Model Coefficients",
       x = "Variables",
       y = "Coefficient Value") +
  theme_minimal()


# Predictions of test data
test_dat_mlr <- read.csv(file.choose(), header=TRUE)

#Separate the Categorical Variables to match train data structure
numer_features <- names(test_dat_mlr)[sapply(test_dat_mlr, is.numeric)] 
categ_features <- test_dat_mlr[, c("type", "location")]

# Create dummy variables for the specified categorical features
dummies_mlr <-  model.matrix(~ type + location - 1, data = test_dat_mlr)

# Combine the numeric features and the dummy variables
test_data_lr <- cbind(test_dat_mlr[, numer_features],dummies_mlr)

# Make predictions
kwality <- round(predict(model_lr, newdata = test_data_lr))

New_wine_data_lr <- data.frame(test_data_lr, kwality)

#Save the Pridicted Data as a CSV file
#write.csv(New_wine_data_lr, "Wine Predicted Data.csv")

# Compare predictions with actual quality
results_lr <- data.frame(Actual_lr = New_wine_data_lr$kwality, Predicted_lr = kwality)
print(results_lr)

# Plotting Actual vs Predicted for Multiple Linear Regression
ggplot(New_wine_data_lr, aes(x = New_wine_data_lr$kwality, y = kwality)) +
    geom_point(color = "blue", alpha = 0.5) +
    geom_abline(slope = 1, intercept = 0, color = "red") +
    labs(title = "Multiple Linear Regression: Actual vs Predicted Quality",
         x = "Actual Wine Quality",
         y = "Predicted Wine Quality") +
    theme_minimal()

# Calculate RMSE
rmse_lr <- sqrt(mean((kwality - New_wine_data_lr$kwality) ^ 2))
cat("Root Mean Squared Error:", rmse_lr, "\n")

# Create confusion matrix
confusion_mat_lr <- confusionMatrix(as.factor(results_lr$Predicted_lr), as.factor(results_lr$Actual_lr))
print(confusion_mat_lr)

# Create a confusion matrix table
cm_table_lr <- as.data.frame(confusion_mat_lr$table)

# Plot confusion matrix
ggplot(data = cm_table_lr, aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "blue") +
  geom_text(aes(label = Freq), vjust = 1) +
  labs(title = "Confusion Matrix", x = "Actual Quality", y = "Predicted Quality") +
  theme_minimal()

# Extract TN, FP, FN, TP
TN_lr <- confusion_mat_lr$table[1, 1]
TP_lr <- confusion_mat_lr$table[2, 2]
FP_lr <- confusion_mat_lr$table[1, 2]
FN_lr <- confusion_mat_lr$table[2, 1]

# Print the values
cat("True Negatives (TN):", TN_lr, "\n")
cat("True Positives (TP):", TP_lr, "\n")
cat("False Positives (FP):", FP_lr, "\n")
cat("False Negatives (FN):", FN_lr, "\n")

# Create a data frame for plotting confusion matrix values
cm_values_lr <- data.frame(
  Type = c("True Negatives", "False Positives", "False Negatives", "True Positives"),
  Count = c(TN_lr, FP_lr, FN_lr, TP_lr)
)

# Plot confusion matrix values
ggplot(cm_values_lr, aes(x = Type, y = Count, fill = Type)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("yellow", "blue", "orange", "skyblue")) +
  labs(title = "Confusion Matrix Values", x = "Type", y = "Count") +
  theme_minimal() +
  theme(legend.position = "none")


```

```{r}

# Compare the RMSE of the two Models I built (Decision Tree & Multiple Linear Regression Model)

# Print RMSE values
cat("RMSE for Multiple Linear Regression Model:", rmse_lr, "\n")
cat("RMSE for Random Forest Model:", rmse_rf, "\n")

# Compare RMSE and pick the better model
if (rmse_lr < rmse_rf) {
  cat("The Linear Regression model is better.\n")
} else if (rmse_rf < rmse_lr) {
  cat("The Random Forest model is better.\n")
} else {
  cat("Both models have the same RMSE.\n")
}

```
```{r}

# Replace actual_values with the actual outcomes in your dataset
predictions_lr <- predict(model_lr, newdata = test_data_lr)
predictions_rf <- predict(rf_model, newdata = test_data_rf)

actual_values_lr <- results_lr$Actual 
actual_values_rf <- results_rf$Actual 

# Calculate accuracy for both models
accuracy_lr <- mean(predictions_lr == actual_values_lr)
accuracy_rf <- mean(predictions_rf == actual_values_rf)

# Print accuracy values
cat("Accuracy for Multiple Linear Regression Model:", accuracy_lr, "\n")
cat("Accuracy for Random Forest Model:", accuracy_rf, "\n")

# Compare accuracy and pick the better model
if (accuracy_lr > accuracy_rf) {
  cat("The Linear Regression model is better.\n")
} else if (accuracy_rf > accuracy_lr) {
  cat("The Random Forest model is better.\n")
} else {
  cat("Both models have the same accuracy.\n")
}


```

################################
#CONCLUSIONS
################################
#1 EDA CONCLUSION
#Summary of Interpretation of ANOVA

#a. Location Effect: The results indicate a highly significant effect of location on wine quality (p < 0.001). This suggests that different locations produce wines of significantly different quality. But the dataset did not show what location. More in-depth work is required.

#b. Type Effect: The effect of wine type on quality is also significant (p < 0.05), indicating that the type of wine does have an impact on its quality, although this effect is not as strong as the location effect. We could not clearly see which of the type of wine had more quality based on the dataset.

#c. Interaction Effect: The interaction between location and type is highly significant (p < 0.001). This means that the effect of wine type on quality varies depending on the location, indicating complex relationships between these factors. As mentioned above, more wine data from California and Texas id required to do more in-depth work 

#d. Conclusion       	
#In conclusion, both location and type significantly influence wine quality, and their interaction suggests that the effect of type is not consistent across different locations. 
#This information can be valuable for winemakers and consumers alike, as it highlights the importance of both geographical and varietal factors in determining wine quality.

#2. CONCLUSION ON THE MODEL BUILT ON THE TRAIN DATA
#a. The comparison result of the models showed that both models predicted good on test data. There is no better models between the models I used according to my analysis.

#b. Since the analysis show that the results are both good, I will then use by subjective opinion to choose the model I will use on the "GIVEN" test data. I prefer MULTIPLE LINEAR REGRESSION.

#c. This is because of the feature importance chosen my the model, which is DENSITY. The feature chosen my RANDOM FOREST did not clearly stated a particular variable, but provided a clue that "location" is the most important feature. I do not know whether it is California or Texas. The dummy used did not explicitly splitted that variable for me to know exactly.

#d. Therefore, I have choosen to use MULTIPLE LINEAR REGRESSION as my model used to predict WINE QUALITY on the given test data of the project.

#e. The trained data was saved as a RDS file and included in my submission. This is used as trianed data for my Shiny App file, named "Wine Quality Prediction App.rmd".

#f. The Test Predicted Wine Quality was saved as a CSV file and included in my submission. This file is named "Wine Predicted Data.csv"




